Housing Market Institutions Drive Race and Ethnicity Differences in Energy Consumption

### 2015-04-01 ConferencePaper/DraftEdwardsWoods.Rnw
editing text, removing junk

The choice of housing, differentiated between renters and owners: owners and those households making payments towards ownership are treated as the same.  Homes are categorized by structure type.   
Housing choices may be linked to fundamental characteristics of the heads of households as a dependent variable that has implications as to utility, and in this respect it is assumed that owning a detached home is of greatest use.  The choice of homes is considered cardinal, where a house is of greatest value and the mobile home is treated as the least appealing choice.  A townhome, row house, in a desirable location such as an urban area, may be a highly desirable good, in general that is not treated as the norm. Structures other than detached houses may be considerably more perishable.  This assumption is made regardless of variations within the desirability of particular detached houses. The choice of purchasing or renting a home is relevant within the structure types.  An individual may choose to forego ownership of a townhouse in favor of renting a house, or forego owning a mobile home in favor of renting an apartment.  Here, both from a utilitarian perspective of the need for one domicile over another is considered as well as prestige that comes from both the structure type as well as what utility may derive from ownership.
	The value of structure type may have implications as to future wealth.  Homeowners tend towards greater wealth over time, related to home ownership itself(Di 2007).  Homeowners have the opportunity to use their wealth for investment or paying off higher interest debts, such as credit cards, by using home equity.  Homeowners have access to a safety net in terms of debt, and they also have the opportunity of reducing the lifetime costs of debts.  Homeowners may also choose to liquidate their homes, or pass on wealth at at or near the end of their lives.  One source found homeowners held more wealth compared to their contemporary counterparts.  The growth of wealth was also appreciably higher(Di 2007). 
The effect of home ownership and structure choice may have implications for future outcomes of the children of the suggested households.  In that respect, it can be said that there is evidence of children having better outcomes, in terms of education and income, than their peers in households that choose to rent (Boehm 2014).  It is plausible that the outcome benefits are coming from a causal relationship with parental characteristics rather than that of the housing choice, and are co-variables, rather than having an internal causal relationship.  While it is beyond the scope of this paper, it is also plausible that housing choice is indicative of neighborhood choice, though it should be stated that there is no assertion on this statement to be found here.
Households choice of structure type, payment method, and location are often very difficult to change based on the history of a household (Darrah 2014).  Households that have tended towards renting have difficulty, or at least choose not to, buy in the future.  Community attitudes and behaviors in relation to housing are often sticky in changing, and this is of importance as related to the outcomes of children; often mirroring the characteristics of their parents.

@@

Variable and Data Description:

RECS data set is the 2009 publication.  12083 observations are included in the data, of which 11395 are included in the struture-tenure model. Combining dwelling type (dwltype) and the binary response variable Ownrent,Dwltype includes six categories as follows: single family detached house, townhouse (including duplex, or row house), small apartment complex, large apartment complex, mobile home. Mobile homes are notable for having considerably fewer observations than other categories. The model is based on the following variables: income, resident ages, education, and ethnicity.  Logged, midpoints, are used for income. Education is broken into six categories describing highest education level completed by the head of the household: elementary, some high school, high school. Some college/trade/vocational school, college graduate and post graduate degree.  The ethnicity variable is constructed from a composite of six dummy variables for head of households broken into head of households one and two.  The result is one variable for each where from one to six: Native American, Asian/Pacific Islander, Black/African American, Hispanic/Latino, White/Caucasian, and Other.  For our purposes head of household #2 is ignored.
	Approximately 2.7 times as many households choose to own their homes compared to renters.  As well we see that the majority of households, more than all other categories combined, live in singe family detached houses (See Appendix).  Behind single family houses, large apartment complexes were the most likely dwelling types chosen.  Notably, apartments in general, were more commonly rented than owned as compared to the opposite for each other dwelling type.
	With respect to respondent education attainment, it can be noted that the percentages of each category (see appendix) do differ from that of what can be expected.  It is entirely likely that this is an example of the survey effect on selection of respondents, i.e. that individuals with higher education attainment are more likely to respond to mailed surveys.  Comparing to a cursory internet search for example, more respondents hold graduate degrees and bachelor’s degrees, and less high school diplomas and below than that percentage constructed for the state of California (PPIC).  It is also of minor interest to note the seventy-nine respondents who claim to live in mobile homes and hold graduate degrees.  While that combination of characteristics is plausible enough without further examination, it does bring into question the honesty of respondent’s self-reported information.  But, by constructing a variable equal to the number of housemates over the age of 65 and tabulating it against just mobile homes within the dwelling type variable, we note that the majority of respondents state that 1346 seniors live in mobile homes compared to 643 mobile homes without senior citizens, which is a considerably different trend than homes largely without seniors in the other dwelling type categories particularly detached houses, bearing in mind that this is relatively and the majority of seniors recorded in the data do not live in mobile homes. This may suggests that seniors are utilizing mobile homes as retirement homes, possibly within retirement communities and that would help to explain the minor perceived discrepancy in the education variable.  It is also commonly held that seniors may be more likely to respond to surveys, much like voting, and this would suggest why there are more respondents in mobile homes than in every other category of dwelling aside from home ownership and small complex ownership.  It can be noted from the RASS survey that, in fact, mobile homes had the highest response rate (RASS).

Lastly, one of our variables, actually two, are count variables that account for the number of children under the age of five and between six and eighteen respectively.  This variable was constructed with the intention of finding whether larger structures, in general, such as houses and townhouses, are chosen more often by individuals requiring more room or who are exhibiting more permanent nesting behavior, as in choosing to purchase a home with the intention of residing there for a long period of time and becoming well established.

@@

Model: Multinomial logit

As a point of clarity, surveys with questions not filled in were coded with 97, to differentiate between responses that were left blank and those where the value coded was not applicable.  For the purpose of this paper 97s were recoded to not and were dropped from the analysis.  
The multinomial logit is first ran with owndwel as the dependent variable and the independent variables as follows: income, educ, race5, age_1, age_2.  The base chosen was that of house renters, represented here as the first category (see appendix).  Here, race5 represents whether the head of the household is white.  Running the multinomial logit with all raises included as independent variables proved problematic, and largely insignificant.  Ages of children were left separate in an attempt to model the difference between households with children of all ages and those with only young children.  Having only young children may represent households in starter homes who are not necessarily well established.  As well households with children over the age of eighteen are essentially ignored and treated as adults.  Another iteration of the model is also down removing the children variables and instead using resident count.
The pseudo r-squared for the first model specified is 0.0912, which low it is important to bear in mind the problematic nature of interpreting pseudo r-squared values.

>>>>

The choice of housing, differentiated between renters and owners: owners and those households making payments towards ownership are treated as the same.  Homes are categorized by structure type.   
Housing choices may be linked to fundamental characteristics of the heads of households as a dependent variable that has implications as to utility, and in this respect it is assumed that owning a detached home is of greatest use.  The choice of homes is considered cardinal, where a house is of greatest value and the mobile home is treated as the least appealing choice.  A townhome, row house, in a desirable location such as an urban area, may be a highly desirable good, in general that is not treated as the norm. Structures other than detached houses may be considerably more perishable.  This assumption is made regardless of variations within the desirability of particular detached houses. The choice of purchasing or renting a home is relevant within the structure types.  An individual may choose to forego ownership of a townhouse in favor of renting a house, or forego owning a mobile home in favor of renting an apartment.  Here, both from a utilitarian perspective of the need for one domicile over another is considered as well as prestige that comes from both the structure type as well as what utility may derive from ownership.
	The value of structure type may have implications as to future wealth.  Homeowners tend towards greater wealth over time, related to home ownership itself(Di 2007).  Homeowners have the opportunity to use their wealth for investment or paying off higher interest debts, such as credit cards, by using home equity.  Homeowners have access to a safety net in terms of debt, and they also have the opportunity of reducing the lifetime costs of debts.  Homeowners may also choose to liquidate their homes, or pass on wealth at at or near the end of their lives.  One source found homeowners held more wealth compared to their contemporary counterparts.  The growth of wealth was also appreciably higher(Di 2007). 
The effect of home ownership and structure choice may have implications for future outcomes of the children of the suggested households.  In that respect, it can be said that there is evidence of children having better outcomes, in terms of education and income, than their peers in households that choose to rent (Boehm 2014).  Households choice of structure type, payment method, and location are often very difficult to change based on the history of a household (Darrah 2014).  Households that have tended towards renting have difficulty, or at least choose not to, buy in the future.  Community attitudes and behaviors in relation to housing are often sticky in changing, and this is of importance as related to the outcomes of children; often mirroring the characteristics of their parents.

@@

Variable and Data Description: 
Residential Energy Consumption Survey national data 2009 (RECS) was utilitized for this portion of the project. 12083 observations are included in the data, of which 11395 are included in the structure-tenure model. Combining dwelling type (dwltype) and the binary response variable ownrent, Dwltype includes six categories as follows: single family detached house, townhouse (including duplex, or row house), small apartment complex, large apartment complex, and mobile home. Mobile homes are notable for having considerably fewer observations than other categories. The model is based on the following variables: income, resident ages, education, rental assistence, and ethnicity and a variable for HUD complaints by state.  Householder race was interacted with income and  HUD complaints.  Logged, midpoints, are used for income. Education is broken into six categories describing highest education level completed by the head of the household: elementary, some high school, high school. Some college, Associate’s, Bachelor’s, Master;s, Professional, and Doctorate degree.  The ethnicity variable is from one to six: Native American, Asian/Pacific Islander, Black/African American, Hispanic/Latino, White/Caucasian, other, and more than one race.  
  Over twice as many households choose to own their homes compared to renters.  As well we see that the majority of households, more than all other categories combined, live in single family detached houses (See Appendix).  Behind single family houses, large apartment complexes were the most likely dwelling types chosen.  Notably, apartments in general, were more commonly rented than owned as compared to the opposite for each other dwelling type.
	With respect to respondent education attainment, percentages of each category (see appendix) do differ from what can be expected.  It is entirely likely that this is an example of the survey effect on selection of respondents, i.e. that individuals with higher education attainment are more likely to respond to mailed surveys.

@@

Model: Multinomial logit
 old info  
The multinomial logit is first ran with owndwel as the dependent variable and the independent variables as follows: income, educ, race5, age_1, age_2.  The base chosen was that of house renters, represented here as the first category (see appendix).  Here, race5 represents whether the head of the household is white.  Running the multinomial logit with all raises included as independent variables proved problematic, and largely insignificant.  Ages of children were left separate in an attempt to model the difference between households with children of all ages and those with only young children.  Having only young children may represent households in starter homes who are not necessarily well established.  As well households with children over the age of eighteen are essentially ignored and treated as adults.  Another iteration of the model is also down removing the children variables and instead using resident count.
The pseudo r-squared for the first model specified is 0.0912, which low it is important to bear in mind the problematic nature of interpreting pseudo r-squared values.
